---
title: "DiffBind Analysis"
output: html_notebook
---

# Imports

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# Load packages
library(DiffBind)
packageVersion("DiffBind")
```

# Sample Sheets
Create sample sheets and comparison contrasts for erythroid ATAC-seq and erythroid RAD21 ChIP-seq

## Erythroid ATAC-seq

```{r}
ery_atac_bam_dir <- "Example_Data/Erythroid_ATAC/BAMs/"
ery_atac_peak_dir <- "Example_Data/Erythroid_ATAC/LanceOtron_Peaks"

ery_atac_bam_dir <- "/ceph/project/Wellcome_Discovery/datashare/towilson/Public_Data/Human/Erythroid/50_Donors_Erythroid/ZEN_T_Wilson_2026/ATAC/hg38/BAMs/"
ery_atac_peak_dir <- "/ceph/project/Wellcome_Discovery/datashare/towilson/Public_Data/Human/Erythroid/50_Donors_Erythroid/ZEN_T_Wilson_2026/ATAC/hg38/LanceOtron_Peaks"

# Set sample name GSM
gsm_ids <- paste0("GSM", 9320314:9320327)

# Sample names
samples <- c(paste0("Don002_ATAC_d13_rep", 1:7),
             paste0("Don003_ATAC_d13_rep", 1:7))

# Erythroid ATAC-seq sample sheet
ery_atac_samples <- data.frame(SampleID = c(paste0("Don002_ATAC_d13_rep", 1:7),
                                            paste0("Don003_ATAC_d13_rep", 1:7)),
                               Tissue = rep("CD34_erythroid", 14),
                               Factor = c(rep("donor_2", 7), 
                                          rep("donor_3", 7)),
                               Condition = rep("healthy", 14),
                               Treatment = rep("bulk", 14),
                               Replicate = c(1:7, 1:7),
                               bamReads = file.path(ery_atac_bam_dir,
                                                    paste0(gsm_ids, "_", samples, ".bam")),
                               Peaks = file.path(ery_atac_peak_dir,
                                                 paste0(gsm_ids, "_", samples, "_L-tron.bed")),
                               PeakCaller = rep("bed", 14))

# Set samples and contrast for donor 2 replicate comparison
ery_don2_samples <- ery_atac_samples[(ery_atac_samples["Factor"] == "donor_2"),][1:6,]
ery_don2_samples["Factor"] = c(rep("group_1", 3), rep("group_2", 3))
ery_don2_contrast <- list(Factor = "group_1")

# Set samples and contrast for donor 2 vs donor 3 replicate comparison
ery_don2_vs_don3_samples <- ery_atac_samples
ery_don2_vs_don3_contrast <- list(Factor = "donor_2")
```

## Erythroid RAD21 ChIP-seq

```{r}
ery_rad21_bam_dir <- "Example_Data/Erythroid_RAD21/BAMs"
ery_rad21_peak_dir <- "Example_Data/Erythroid_RAD21/LanceOtron_Peaks"

# Erythroid RAD21 ChIP-seq sample sheet
ery_rad21_samples <- data.frame(SampleID = c(paste0("Don001_RAD21_d13_rep", 1:3),
                                             paste0("Don002_RAD21_d13_rep", 1:3),
                                             "Don030_RAD21_d13_rep_1",
                                             "Don030_RAD21_d13_rep_3"),
                                Tissue = rep("CD34_erythroid", 8),
                                Factor = c(rep("donor_1", 3),
                                           rep("donor_2", 3),
                                           rep("donor_30", 2)),
                                Condition = rep("healthy", 8),
                                Treatment = rep("bulk", 8),
                                Replicate = c(1:3, 1:3, 1, 3),
                                bamReads = c(file.path(ery_rad21_bam_dir, paste0("Don001_RAD21_d13_rep", 1:3, ".bam")),
                                             file.path(ery_rad21_bam_dir, paste0("Don002_RAD21_d13_rep", 1:3, ".bam")),
                                             file.path(ery_rad21_bam_dir, "Don030_RAD21_d13_rep1.bam"),
                                             file.path(ery_rad21_bam_dir, "Don030_RAD21_d13_rep3.bam")),
                                Peaks = c(file.path(ery_rad21_peak_dir, paste0("Don001_RAD21_d13_rep", 1:3, ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don002_RAD21_d13_rep", 1:3, ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don030_RAD21_d13_rep1", ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don030_RAD21_d13_rep3", ".bed"))),
                            PeakCaller = rep("bed", 8))
```

# Extract LIB, RLE and TMM Scaling Factors
Extract scaling factors to rescale bigWigs by full library size (LIB), RLE and TMM normalisation

```{r}
# Set sample sheets per dataset
sample_sheets <- list("Erythroid_ATAC" = ery_atac_samples, "Erythroid_RAD21" = ery_rad21_samples)
# Autosomal and sex chromosomes
chromosomes <- c(paste0("chr", 1:22), "chrX", "chrY")
# Normalisation methods to test
norm_methods <- c(DBA_NORM_LIB, DBA_NORM_RLE, DBA_NORM_TMM)

scaling_factors <- list()

for (data in c("Erythroid_ATAC", "Erythroid_RAD21")) {
  # Create DBA object
  dba_object <- dba(sampleSheet = sample_sheets[[data]])
  
  # Find chromosomes to remove
  peakset <- dba.peakset(dba_object, bRetrieve = TRUE)
  greylist <- peakset[!(seqnames(peakset) %in% chromosomes)]
  
  # Apply ENCODE blacklist and keep only standard chromosomes
  dba_object <- dba.blacklist(dba_object, blacklist = DBA_BLACKLIST_HG38, greylist = greylist)
  
  # Re-center peaks and count reads
  dba_object <- dba.count(dba_object)
  
  # Normalise by each method
  for (norm in norm_methods) {
    dba_object <- dba.normalize(dba_object, normalize = norm)
    # Extract the scaling factors
    norm_info <- dba.normalize(dba_object, bRetrieve = TRUE)
    scalers <- norm_info$norm.factors
    names(scalers) <- dba_object$samples$SampleID
    scaling_factors[[data]][[norm]] <- scalers
  }
}
```

```{r}
# View scaling factors per dataset and normalisation method
scaling_factors
```

# Differential Accessibility Analysis

## ZEN Scaling Factors

```{r}
# Scaling factors per sample for ZEN normalisation
ery_atac_zen <- c(Don002_ATAC_d13_rep1 = 0.2795583724430078,
                  Don002_ATAC_d13_rep2 = 0.24549477454191823,
                  Don002_ATAC_d13_rep3 = 0.1651864697590268,
                  Don002_ATAC_d13_rep4 = 0.09541567508135224,
                  Don002_ATAC_d13_rep5 = 0.16076437035969782,
                  Don002_ATAC_d13_rep6 = 0.1551683098985989,
                  Don002_ATAC_d13_rep7 = 0.143536842576444,
                  Don003_ATAC_d13_rep1 = 0.2582363354917962,
                  Don003_ATAC_d13_rep2 = 0.25053836607363844,
                  Don003_ATAC_d13_rep3 = 0.17587857961294,
                  Don003_ATAC_d13_rep4 = 0.11435196335506045,
                  Don003_ATAC_d13_rep5 = 0.10978229659749278,
                  Don003_ATAC_d13_rep6 = 0.1178921511723179,
                  Don003_ATAC_d13_rep7 = 0.0877359143756193)
```

## Run DiffBind

```{r}
# Create DBA object
sheet <- ery_don2_vs_don3_samples
contrast <- ery_don2_vs_don3_contrast
  
dba_object <- dba(sampleSheet = sheet)
dba_object
```

```{r}
# Clustering by peak calls
dba.plotHeatmap(dba_object)
```

```{r}
# Find chromosomes to remove
chromosomes <- c(paste0("chr", 1:22), "chrX", "chrY")
peakset <- dba.peakset(dba_object, bRetrieve = TRUE)
greylist <- peakset[!(seqnames(peakset) %in% chromosomes)]

# Apply ENCODE blacklist and keep only standard chromosomes
dba_object <- dba.blacklist(dba_object, blacklist = DBA_BLACKLIST_HG38, greylist = greylist)

# Re-center peaks and count reads
dba_object <- dba.count(dba_object)
dba_object
```

```{r}
# DBA_NORM_LIB (default), DBA_NORM_RLE (DESeq2), DBA_NORM_TMM (edgeR) or custom vector (ZEN)
norm_method <- "ZEN"
scalars <- ery_atac_zen
  
if (norm_method == "ZEN") {
  # Select scaling factors for the specific replicates
  norm_sub <- scalars[names(scalars) %in% sheet$SampleID]
} else {
  norm_sub <- norm_method
}

dba_object <- dba.normalize(dba_object, normalize = norm_sub)
norm_info <- dba.normalize(dba_object, bRetrieve = TRUE)
# View the scaling factors
norm_info
```

```{r}
# Clustering by count scores
plot(dba_object)
```

```{r}
# Set the contrast
dba_object <- dba.contrast(dba_object, reorderMeta = contrast)
dba_object <- dba.analyze(dba_object)
dba.show(dba_object, bContrasts = TRUE)
```

```{r}
plot(dba_object, contrast = 1)
```

```{r}
results <- dba.report(dba_object, contrast = 1)
results
```

```{r}
output_file <- paste0("Erythroid_ATAC/DiffBind_Erythroid_Donor2_vs_Donor3_", norm_method, ".csv")

results_df <- as(results, "data.frame")
results_df <- results_df[c("seqnames", "start", "end", "Fold", "p-value", "FDR")]
names(results_df)[names(results_df) == "seqnames"] <- "chrom"
print(paste("Significant:", nrow(results_df[results_df$FDR < 0.05, ])))
write.csv(results_df, output_file, row.names = FALSE)
results_df
```