---
title: "DiffBind Analysis"
output: html_notebook
---

# Imports

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# Load packages
library(DiffBind)
```

# Sample Sheets
Create sample sheets and comparison contrasts for erythroid ATAC-seq and erythroid RAD21 ChIP-seq

## Erythroid ATAC-seq

```{r}
ery_atac_bam_dir <- "Example_Data/Erythroid_ATAC/BAMs"
ery_atac_peak_dir <- ""

# Erythroid ATAC-seq sample sheet
ery_atac_samples <- data.frame(SampleID = c(paste0("Don002_ATAC_d13_rep", 1:7),
                                            paste0("Don003_ATAC_d13_rep", 1:7)),
                               Tissue = rep("CD34_erythroid", 14),
                               Factor = c(rep("donor_2", 7), 
                                          rep("donor_3", 7)),
                               Condition = rep("healthy", 14),
                               Treatment = rep("bulk", 14),
                               Replicate = c(1:7, 1:7),
                               bamReads = c(file.path(paste0(ery_atac_bam_dir, "Don002_ATAC_d13_rep", 1:7, ".bam")),
                                            file.path(paste0(ery_atac_bam_dir, "Don003_ATAC_d13_rep", 1:7, ".bam"))),
                               Peaks = c(file.path(ery_atac_peak_dir, paste0("Don002_ATAC_d13_rep", 1:7, ".bed")),
                                         file.path(ery_atac_peak_dir, paste0("Don003_ATAC_d13_rep", 1:7, ".bed"))),
                               PeakCaller = rep("bed", 14))

# Set samples and contrast for donor 2 replicate comparison
ery_don2_samples <- ery_atac_samples[(ery_atac_samples["Factor"] == "donor_2"),][1:6,]
ery_don2_samples["Factor"] = c(rep("group_1", 3), rep("group_2", 3))
ery_don2_contrast <- list(Factor = "group_1")

# Set samples and contrast for donor 2 vs donor 3 replicate comparison
ery_don2_vs_don3_samples <- ery_atac_samples[c(1,2,3,8,9,10),]
ery_don2_vs_don3_contrast <- list(Factor = "donor_2")
```

## Erythroid RAD21 ChIP-seq

```{r}
ery_rad21_bam_dir <- "Example_Data/Erythroid_RAD21/BAMs"
ery_rad21_peak_dir <- ""

# Erythroid RAD21 ChIP-seq sample sheet
ery_rad21_samples <- data.frame(SampleID = c(paste0("Don001_RAD21_d13_rep", 1:3),
                                             paste0("Don002_RAD21_d13_rep", 1:3),
                                             "Don030_RAD21_d13_rep_1",
                                             "Don030_RAD21_d13_rep_3"),
                                Tissue = rep("CD34_erythroid", 8),
                                Factor = c(rep("donor_1", 3),
                                           rep("donor_2", 3),
                                           rep("donor_30", 2)),
                                Condition = rep("healthy", 8),
                                Treatment = rep("bulk", 8),
                                Replicate = c(1:3, 1:3, 1, 3),
                                bamReads = c(file.path(ery_rad21_bam_dir, paste0("Don001_RAD21_d13_rep", 1:3, ".bam")),
                                             file.path(ery_rad21_bam_dir, paste0("Don002_RAD21_d13_rep", 1:3, ".bam")),
                                             file.path(ery_rad21_bam_dir, "Don030_RAD21_d13_rep1.bam"),
                                             file.path(ery_rad21_bam_dir, "Don030_RAD21_d13_rep3.bam")),
                                Peaks = c(file.path(ery_rad21_peak_dir, paste0("Don001_RAD21_d13_rep", 1:3, ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don002_RAD21_d13_rep", 1:3, ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don030_RAD21_d13_rep1", ".bed")),
                                          file.path(ery_rad21_peak_dir, paste0("Don030_RAD21_d13_rep3", ".bed"))),
                            PeakCaller = rep("bed", 8))

# Set the contrast
contrast <- list(Factor = "donor_1")
```

# Extract LIB, RLE and TMM Scaling Factors

Extract scaling factors to rescale bigWigs by full library size (LIB), RLE and TMM normalisation

```{r}
# Set sample sheets per dataset
sample_sheets <- list("Erythroid_ATAC" = ery_atac_samples, "Erythroid_RAD21" = ery_rad21_samples)
# Autosomal and sex chromosomes
chromosomes <- c(paste0("chr", 1:22), "chrX", "chrY")
# Normalisation methods to test
norm_methods <- c(DBA_NORM_LIB, DBA_NORM_RLE, DBA_NORM_TMM)

scaling_factors <- list()

for (data in c("Erythroid_ATAC", "Erythroid_RAD21")) {
  # Create DBA object
  dba_object <- dba(sampleSheet = sample_sheets[[data]])
  
  # Find chromosomes to remove
  peakset <- dba.peakset(dba_object, bRetrieve = TRUE)
  greylist <- peakset[!(seqnames(peakset) %in% chromosomes)]
  
  # Apply ENCODE blacklist and keep only standard chromosomes
  dba_object <- dba.blacklist(dba_object, blacklist = DBA_BLACKLIST_HG38, greylist = greylist)
  
  # Re-center peaks and count reads
  dba_object <- dba.count(dba_object)
  
  # Normalise by each method
  for (norm in norm_methods) {
    dba_object <- dba.normalize(dba_object, normalize = norm)
    # Extract the scaling factors
    norm_info <- dba.normalize(dba_object, bRetrieve = TRUE)
    scalers <- norm_info$norm.factors
    names(scalers) <- dba_object$samples$SampleID
    scaling_factors[[data]][[norm]] <- scalers
  }
}
```
```{r}
# View scaling factors per dataset and normalisation method
scaling_factors
```

# Differential Accessibility Analysis

## ZEN Scaling Factors

```{r}
# Scaling factors per sample for ZEN normalisation
ery_atac_zen <- c(Don002_ATAC_d13_rep1 = 0.2100296353671081,
                  Don002_ATAC_d13_rep2 = 0.1795845107811403,
                  Don002_ATAC_d13_rep3 = 0.10619528895735125,
                  Don002_ATAC_d13_rep4 = 0.0713442611529159,
                  Don002_ATAC_d13_rep5 = 0.12978024048757197,
                  Don002_ATAC_d13_rep6 = 0.12575312989866483,
                  Don002_ATAC_d13_rep7 = 0.10672348945554712,
                  Don003_ATAC_d13_rep1 = 0.1978762445037181,
                  Don003_ATAC_d13_rep2 = 0.17489888374729184,
                  Don003_ATAC_d13_rep3 = 0.11213534059402934,
                  Don003_ATAC_d13_rep4 = 0.07863437918851453,
                  Don003_ATAC_d13_rep5 = 0.08186895342833504,
                  Don003_ATAC_d13_rep6 = 0.09581304059400693,
                  Don003_ATAC_d13_rep7 = 0.06246986948347924)
```

## Run DiffBind

```{r}
# Create DBA object
dba_object <- dba(sampleSheet = ery_bulk_don2_samples)
dba_object
```

```{r}
# Clustering by peak calls
dba.plotHeatmap(dba_object)
```

```{r}
# Find chromosomes to remove
chromosomes <- c(paste0("chr", 1:22), "chrX", "chrY")
peakset <- dba.peakset(dba_object, bRetrieve = TRUE)
greylist <- peakset[!(seqnames(peakset) %in% chromosomes)]

# Apply ENCODE blacklist and keep only standard chromosomes
dba_object <- dba.blacklist(dba_object, blacklist = DBA_BLACKLIST_HG38, greylist = greylist)

# Re-center peaks and count reads
dba_object <- dba.count(dba_object)
dba_object
```

```{r}
# DBA_NORM_LIB (default), DBA_NORM_RLE (DESeq2), DBA_NORM_TMM (edgeR) or custom vector (ZEN)
dba_object <- dba.normalize(dba_object, normalize = DBA_NORM_RLE)
norm_info <- dba.normalize(dba_object, bRetrieve = TRUE)
# View the scaling factors
norm_info
```

```{r}
# Clustering by count scores
plot(dba_object)
```

```{r}
# Set the contrast
dba_object <- dba.contrast(dba_object, reorderMeta = contrast)
dba_object <- dba.analyze(dba_object)
dba.show(dba_object, bContrasts = TRUE)
```

```{r}
plot(dba_object, contrast = 1)
```

```{r}
results <- dba.report(dba_object, contrast = 1)
results
```

```{r}
output_file <- "Erythroid_ATAC/DiffBind_Erythroid_Donor2_vs_Donor3_RLE.csv"

results_df <- as(results, "data.frame")
results_df <- results_df[c("seqnames", "start", "end", "Fold", "p-value", "FDR")]
names(results_df)[names(results_df) == "seqnames"] <- "chrom"
print(paste("Significant:", nrow(results_df[results_df$FDR < 0.05, ])))
write.csv(results_df, output_file, row.names = FALSE)
results_df
```